---
title: The `doBy` package
author:
  - name: Søren Højsgaard
    affiliation: Department of Mathematical Sciences, Aalborg University, Denmark
    address:
    - line 1
    - line 2
    email:  sorenh@math.aau.dk
abstract: >
  The `doBy` is one of several general utility packages on CRAN. 
  An abstract of less than 150 words.
preamble: |
  % Any extra LaTeX you need in the preamble
output: rticles::rjournal_article
---

```{r, echo=FALSE}
library(doBy)
library(magrittr)
options("digits"=4)
```


## Introduction

The `doBy` package [@doby] appeared on CRAN [@CRAN] in 2006 and, much
to our surprise, the package is still being used.  The package
originally grew out of a need to calculate groupwise summary
statistics (much in the spirit of `PROC SUMMARY` of the `SAS` system,
[@procsummary]). The name comes from `doing` som computations when
data is stratified `by` the value of some variables.  Today the
package contains many different utilities. In this paper we focus 1)
on these "doing by" functions, 2) on functions related to linears
estimates and contrasts and 3) on some of the miscellaneous functions
in the package.



## Functions related to groupwise computations

### A working dataset

The `CO2` data frame comes  from an
experiment on the cold tolerance of the grass species _Echinochloa
crus-galli_.
To limit the amount of output we modify names and levels of variables
as follows

```{r}
data(CO2)
CO2 <- within(CO2, {
    Treat <- Treatment
    Treatment <- NULL
    levels(Treat) <- c("nchil","chil")
    levels(Type)  <- c("Que","Mis")
})
CO2 <- subset(CO2, Plant %in% c("Qn1", "Qc1", "Mn1", "Mc1"))
CO2
```

### The `summaryBy` function

The `summaryBy` function is used for calculating quantities like __the
mean and variance of numerical variables `x` and `y` for each
combination of two factors `A` and `B$`__.  Notice: A functionality
similar to `summaryBy` is provided by `aggregate` from base R. [SH:
DUMT at variabel skal navngives; det skal den ikke i aggregate].

```{r}
myfun1 <- function(x){c(m=mean(x), s=sd(x))}
summaryBy(cbind(conc, uptake, lu=log(uptake)) ~ Plant, 
          data=CO2, FUN=myfun1)
```

Instead of formula we may specify a list containing the left hand side
and the right hand side of a formula (This is a feature of
`summaryBy` and it does not work with `aggregate`.). 

```{r}
## FIXME: Will fail because of log(uptake) 
summaryBy(list(c("conc", "uptake", "lu=log(uptake)"), "Plant"), data=CO2, FUN=myfun1)
```

Various convenient abbreviations are available, e.g. the following, where 
left hand side dot refers to "all numeric variables" while the
right hand side dot refers to "all factor variables".

```{r}
summaryBy(.~., data=CO2, FUN=myfun1)
## same as summaryBy(list(c("."), c(".")), data=CO2, FUN=myfun1)
```

### The `orderBy` function

Ordering (or sorting) a data frame is possible with the `orderBy`
function.  Suppose we want to order the rows of the the `CO2` data by
increasing values of `conc` and decreasing value of `uptake` (within `code`):
```{r}
x1 <- orderBy(~ conc - uptake, data=CO2)
head(x1)
```

An equivalent  form is:
```{r}
orderBy(c("conc", "-uptake"), data=CO2) %>% head
```

### The `splitBy` function

Suppose we want to split `CO2` into a list of dataframes:

```{r}
x1 <- splitBy(~ Plant + Type, data=CO2)
x1
```

The result is a list:
```{r}
lapply(x1, head, 2)
```

### The `subsetBy` function


Suppose we want to select those rows within each treatment for which  the
uptake is larger than 75% quantial of uptake (within the treatment). This
is achieved by:

```{r}
x2 <- subsetBy(~Treat, subset=uptake > quantile(uptake, prob=0.75), data=CO2)
x2
```

### The `transformBy` function

The `transformBy` function is analogous to the `transform`
function except that it works within groups. For example:

```{r}
x <- transformBy(~Treat, data=CO2, 
                 minU=min(uptake), maxU=max(uptake),
                 range = diff(range(uptake)))
head(x)
```

## The `lmBy` function

```{r}
m <- lmBy(uptake ~ conc | Treat, data=CO2)
lapply(m, function(z) coef(summary(z)))
```



## Functions related linear estimates and contrasts

A linear function of a $p$--dimensional parameter vector $\beta$ has
the form
\begin{displaymath}
  C=L\beta
\end{displaymath}
where $L$ is a $q\times p$ matrix which we call the `Linear Estimate Matrix` or 
simply `LE-matrix`.  The corresponding
linear estimate is $\hat C = L \hat \beta$.  A linear hypothesis has
the form $H_0: L\beta=m$ for some $q$ dimensional vector $m$.


### A working dataset

The response is the length of odontoblasts cells (cells responsible for
tooth growth) in 60 guinea pigs.  Each animal received one of
three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of
two delivery methods, (orange juice (coded as OJ) or ascorbic acid 
(a form of vitamin C and (coded as VC)).

```{r}
ToothGrowth$dose <- factor(ToothGrowth$dose)
head(ToothGrowth, 4)
```

### Computing linear estimates

Focus on additive model:
```{r}
tooth1 <- lm(len ~ dose + supp, data=ToothGrowth)
```


Consider computing the estimated length for each dose of orange juice (OJ):
One option: Construct the LE--matrix $L$ directly and then invoke `linest`:

```{r}
L <- matrix(c(1, 0, 0, 0, 
              1, 1, 0, 0,
              1, 0, 1, 0), nrow=3, byrow=T)
c1 <- linest(tooth1, L)
c1
```

We can do:

```{r}
summary(c1)
coef(c1)
confint(c1)
```


The matrix $L$ can be generated as follows:
```{r}
L <- LE_matrix(tooth1, effect="dose", at=list(supp="OJ"))
L
```


There are various alternatives:
```{r}
c1 <- esticon(tooth1, L)
c1
```

Notice: `esticon` has been in the `doBy` package for many years;
`linest` is a newer addition.  Yet another
alternative in this case is to generate a new data frame and then
invoke predict (but this approach is not generally applicable, see later):

```{r}
nd <- data.frame(dose=c('0.5', '1', '2'), supp='OJ')
nd
predict(tooth1, newdata=nd)
```


### Least-squares means (LS--means)

A related question could be: What is the estimated length for each
dose if we ignore the source of vitamin C (i.e.\ whether it is OJ or
VC). One approach would be to fit a model in which source does not appear:

```{r}
tooth0 <- update(tooth1, . ~ . - supp)
L0 <- LE_matrix(tooth0, effect="dose")
L0
linest(tooth0, L=L0)
```
<< >>=
@

An alternative would be to stick to the original model but compute the
estimate for an ``average vitamin C source''. That would correspond to
giving weight $1/2$ to each of the two vitamin C source
parameters. However, as one of the parameters is already set to zero
to obtain identifiability, we obtain the LE--matrix $L$ as

```{r}
L1 <- matrix(c(1, 0, 0, 0.5, 
               1, 1, 0, 0.5,
               1, 0, 1, 0.5), nrow=3, byrow=T)
linest(tooth1, L=L1)
```

Such a particular linear estimate is sometimes called a least-squares
mean or an LSmean or a marginal mean. Notice that the parameter
estimates under the two approaches are identical. This is is because
data is balanced: There are $10$ observations per supplementation
type. Had data not been balanced, the estimates would in general have been different. 

Notice: One may generate $L$ automatically with
```{r}
L1 <- LE_matrix(tooth1, effect="dose")
L1
```

Notice: One may obtain the LSmean directly as:

```{r}
LSmeans(tooth1, effect="dose")
```

%% \subsection{Additive model}
%% \label{sec:additive-model}

%% Returning to the \verb|ToothGrowth| data, orange juice and ascorbic
%% acid are just two of many ways of supplying vitamin C (citrus and lime
%% juice would be two alternatives). Here one can therefore argue, that
%% it would make sense to estimate the the effect for each dose for an
%% ``average vitamin C source'':

%% << >>= 
%% LSmeans(tooth1, effect="dose")
%% @

which is the same as

```{r}
L <- LE_matrix(tooth1, effect="dose")
le <- linest(tooth1, L=L)
coef(le)
```

%% \subsection{Interaction model}
%% \label{sec:interaction-model}


For a model with interactions, the LSmeans are
```{r}
LSmeans(tooth2, effect="dose")
```

In this case, the LE--matrix is
```{r}
L <- LE_matrix(tooth2, effect="dose")
t(L)

```

### Using the `at=` argument


Consider random regression model:
<<>>=
library(lme4)
chick <- lmer(weight ~ Time * Diet + (0 + Time | Chick), 
           data=ChickWeight)
coef(summary(chick))
@ %def


The LE--matrix for \cc{Diet} becomes:
<<>>=
L <- LE_matrix(chick, effect="Diet")
t(L)
@ %def

The value of \cc{Time} is by default taken to be the average of that
variable. Hence the \lsmeans\ is the predicted weight for each diet at
that specific point of time. We can consider other points of time with
<<>>=
K1 <- LE_matrix(chick, effect="Diet", at=list(Time=1))
t(K1)
@ %def

The \lsmeans\ for the intercepts is the predictions at
\cc{Time=0}. The \lsmeans\ for the slopes becomes
<<>>=
K0 <- LE_matrix(chick, effect="Diet", at=list(Time=0))
t(K1 - K0)
linest(chick, L=K1-K0)
@ %def

We can cook up our own function for comparing trends:

<<>>=
LSmeans_trend <- function(object, effect, trend){
    L <- LE_matrix(object, effect=effect, at=as.list(setNames(1, trend))) -
        LE_matrix(object, effect=effect, at=as.list(setNames(0, trend)))
    linest(object, L=L)
}
LSmeans_trend(chick, effect="Diet", trend="Time")
@ %def

\section{Using (transformed) covariates}
\label{sec:using-covariates}

Consider the following subset of the \code{CO2} dataset:

<<>>=
data(CO2)
CO2 <- transform(CO2, Treat=Treatment, Treatment=NULL)
levels(CO2$Treat) <- c("nchil","chil")
levels(CO2$Type) <- c("Que","Mis")
ftable(xtabs( ~ Plant + Type + Treat, data=CO2), col.vars=2:3)
@ %def

<<co2-fig, fig.cap="CO2 data">>=
qplot(x=log(conc), y=uptake, data=CO2, color=Treat, facets=~Type)
@ %def


Below, the covariate \code{conc} is fixed at the average value:
<<>>=
co2.lm1 <- lm(uptake ~ conc + Type + Treat, data=CO2)
LSmeans(co2.lm1, effect="Treat")
@ %def

If we use \code{log(conc)} instead we will get an error when
calculating LS--means:
<<eval=F>>=
co2.lm <- lm(uptake ~ log(conc) + Type + Treat, data=CO2)
LSmeans(co2.lm, effect="Treat")
@ %def

In this case one can do
<<>>=
co2.lm2 <- lm(uptake ~ log.conc + Type + Treat,
             data=transform(CO2, log.conc=log(conc)))
LSmeans(co2.lm2, effect="Treat")
@ %def

This also highlights what is computed: The average of the log of
\code{conc}; not the log of the average of \code{conc}.

In a similar spirit consider

<<>>=
co2.lm3 <- lm(uptake ~ conc + I(conc^2) + Type + Treat, data=CO2)
LSmeans(co2.lm3, effect="Treat")
@ %def

Above \verb'I(conc^2)' is the average of the squared values of
\code{conc}; not the  square of the average of
\code{conc}, cfr.\ the following.

<<>>=
co2.lm4 <- lm(uptake ~ conc + conc2 + Type + Treat, data=
              transform(CO2, conc2=conc^2))
LSmeans(co2.lm4, effect="Treat")
@ %def

If we want to evaluate the LS--means at \code{conc=10} then we can do:
<<>>=
LSmeans(co2.lm4, effect="Treat", at=list(conc=10, conc2=100))
@ %def











## Acknowledgements

Credit is due to Dennis Chabot, Gabor Grothendieck, Paul Murrell, Jim
Robison-Cox and Erik Jørgensen for reporting various bugs and
making various suggestions to the functionality in the `doBy`
package.


## Summary

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).

\bibliography{RJreferences}






<!-- ```{r} -->

<!-- mtcars <- within(mtcars, { -->
<!--    vs <- factor(vs, labels = c("V", "S")) -->
<!--    am <- factor(am, labels = c("auto", "man")) -->
<!-- }) -->
<!-- mtcars$drat <- mtcars$disp <- mtcars$wt <- mtcars$carb <- mtcars$gear <- NULL -->
<!-- mtcars <- subset(mtcars, cyl < 8) -->
<!-- mtcars %>% head -->
<!-- ``` -->





<!-- ```{r} -->
<!-- library(doBy) -->
<!-- summaryBy(cbind(mpg, qsec) ~ cyl + vs, data=mtcars) -->

<!-- summaryBy(list(c("mpg", "qsec"), c("cyl", "vs")), data=mtcars) -->

<!-- summaryBy(. ~ cyl + vs, data=mtcars) -->

<!-- summaryBy(. ~ ., data=mtcars) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ss <- splitBy(~ vs, data=mtcars) -->
<!-- ss -->
<!-- ss$V -->
<!-- ``` -->


<!-- Introductory section which may include references in parentheses -->
<!-- [@R], or cite a reference such as @R in the text. -->

<!-- This section may contain a figure such as Figure \ref{fig:Rlogo}. -->

<!-- ```{r, Rlogo, echo=FALSE, fig.cap='The logo of R.', out.width='2in', fig.align='center', fig.pos='htbp'} -->
<!-- knitr::include_graphics('Rlogo.pdf') -->
<!-- ``` -->

<!-- There will likely be several sections, perhaps including code snippets, such as: -->

<!-- ```{r} -->
<!-- x <- 1:10 -->
<!-- x -->
<!-- ``` -->
