---
title: "Plotting classification regions for LDA and QDA models"
author: "Søren Højsgaard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting classification regions for LDA and QDA models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The function `plot_class_region()` displays classification regions for models
fitted with `MASS::lda()` or `MASS::qda()`.

Two different constructions are available:

- `type = "refit"` fits a new model using only the two predictors shown in a
  plot.
- `type = "slice"` retains the original fitted model and holds all predictors
  not shown in the plot fixed at representative values.

The default is `type = "refit"`.

```{r packages}
library(doBy)
library(MASS)
```

## A single predictor pair

We begin by fitting an LDA model to the `iris` data.

```{r fit}
fit <- lda(
  Species ~ Sepal.Length + Sepal.Width + Petal.Length,
  data = iris
)
```

A classification-region plot for two predictors is obtained by specifying
their names:

```{r one-pair}
plot_class_region(
  fit,
  vars = c("Sepal.Length", "Sepal.Width"),
  data = iris
)
```

Predictors may also be selected by their positions among the predictors in the
model formula:

```{r one-pair-indices}
plot_class_region(
  fit,
  vars = c(1, 3),
  data = iris
)
```

## Refit versus slice

With `type = "refit"`, the model is refitted using only the two displayed
predictors. Predictors not shown in the plot do not enter the refitted model.

```{r refit}
plot_class_region(
  fit,
  vars = c("Sepal.Length", "Petal.Length"),
  data = iris,
  type = "refit"
)
```

With `type = "slice"`, predictions are instead obtained from the original
three-predictor model. The remaining predictor is held fixed, by default at
its mean if it is numeric.

```{r slice}
plot_class_region(
  fit,
  vars = c("Sepal.Length", "Petal.Length"),
  data = iris,
  type = "slice"
)
```

The fixed value can be specified explicitly through `at`:

```{r slice-at}
plot_class_region(
  fit,
  vars = c("Sepal.Length", "Petal.Length"),
  data = iris,
  type = "slice",
  at = list(Sepal.Width = 3)
)
```

A slice is conditional on the chosen fixed values. It should therefore not be
interpreted as a refitted two-predictor model.

## Several predictor pairs

When more than two predictors are supplied, plots are produced for all
pairwise combinations.

```{r several-pairs}
plot_class_region(
  fit,
  vars = c(1, 2, 3),
  data = iris,
  ncol = 3
)
```

The function invisibly returns the individual `ggplot` objects. Setting
`plot = FALSE` suppresses the automatic display and makes it convenient to
arrange or modify the plots afterwards.

```{r return-plots}
plots <- plot_class_region(
  fit,
  vars = c(1, 2, 3),
  data = iris,
  plot = FALSE
)

names(plots)
```

For example, the plots can be arranged in one row with `patchwork`:

```{r arrange-plots, eval=FALSE}
patchwork::wrap_plots(plots, nrow = 1)
```

or modified individually before they are combined:

```{r modify-plots, eval=FALSE}
plots[[1]] <- plots[[1]] + ggplot2::theme_minimal()
patchwork::wrap_plots(plots, nrow = 1)
```

## QDA models

The same interface can be used for quadratic discriminant analysis.

```{r qda}
fit_qda <- qda(
  Species ~ Sepal.Length + Sepal.Width + Petal.Length,
  data = iris
)

plot_class_region(
  fit_qda,
  vars = c("Sepal.Length", "Petal.Length"),
  data = iris
)
```

## Notes

The function currently supports numeric, untransformed main-effect predictors.
Interactions, polynomial terms, and transformed predictors are not supported
as plotting variables.

In a slice plot, the displayed observations retain their actual values for
predictors not shown on the axes. Consequently, the observed points do not
necessarily lie in exactly the same slice as the displayed classification
regions.
